---
title: "GDM_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GDM_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Generalized dissimilarity modeling (GDM)

```{r setup, warning = FALSE, message = FALSE}
library(landgen)
library(gdm)
library(here)
library(tidyverse)
library(raster)
library(rgdal)
library(readr)
```

Generalized dissimilarity modeling is a matrix regression method in which explanatory variables (in our case, genetic data, in the form of a distance matrix) is regressed against a response matrix (environmental variables for sites from which samples were obtained). A GDM calculates the compositional dissimilarity between pairs of sites, and importantly takes into account the fact that genetic variation varies nonlinearly across an environmental gradient.

For additional information on GDMs, please see [Ferrier et al. 2007](https://onlinelibrary.wiley.com/doi/10.1111/j.1472-4642.2007.00341.x) for a description of its basic use in estimating patterns of beta diversity, [Freedman et al. 2010](https://onlinelibrary.wiley.com/doi/10.1111/j.1365-294X.2010.04684.x) for a classic example of its use, and [Fitzpatrick & Keller 2015](https://onlinelibrary.wiley.com/doi/abs/10.1111/ele.12376#:~:text=Community%2Dlevel%20modelling%20of%20genomic,assessments%20of%20climate%20change%20vulnerability.) for a perspective piece on its applications. Finally, our code primarily uses the gdm package which has excellent documentation (see [here](https://cran.r-project.org/web/packages/gdm/gdm.pdf)).

There is one main function to perform a GDM analysis: `gdm_do_everything()`. This function runs the GDM (using the `gdm()` function within the gdm package), and allows a user to run a GDM with all variables, or with model selection to choose the best-supported variables. This function produces:

1.    Information on the final model
2.    Coefficients for predictor variables
3.    A PCA RGB map

There are a few assumptions built within this function that the user must be aware of: (1) the coords and gendist files MUST have the same ordering of individuals; there isn't a check for this, and (2) this function assumes individual-based sampling and that each individual is a sampling site. If this isn't true, TODO COMPLETE THIS

2.  `gdm_coeffs()` - Get coefficients from models to determine support

3.  `gdm_map()` - Create a map given your model

4.  `gdm_plot_isplines()` - Plot isplines

## Read in and process data files

Running a GDM requires three data files for input: a genetic distance matrix (the `gendist` argument), coordinates for samples (the `coords` argument), and environmental layers on which to run the GDM (the `envlayers` argument).

```{r data}
# Load genetic dist matrix and coordinates for 53 inds, and three environmental layers for test dataset
load(here("data", "gendist.RDA"))
load(here("data", "liz_coords.RDA"))
load(here("data", "CA_env.RDA"))

# Our code assumes that the first column is longitude and second is latitude; check this:
head(liz_coords)
# Also, our code assumes that sample IDs from gendist and coords are the same order; be sure to check this before moving forward!
```

## Run GDM

Given that GDM is a regression, the full model (i.e., including all predictor variables) will include all environmental layers in addition to geographic distance, which is also considered a predictor. Thus, in this example, the maximum number of variables you can end up with that are significant is four (three enviro PCs + geographic distance).

### GDM with all variables

Let's first run a full GDM model (i.e., including all four variables), specified using the `model` argument. If you have extracted environmental values for each sampling coordinate, this must be specified using the `env` argument, and if genetic distances are not bounded by 0-1, they must be scaled using the `scale` argument. Keep in mind that the `nPerm` argument is only for model selection (see below) and so will not be used in this case.

```{r all vars}
mod_full <- gdm_do_everything(gendist, liz_coords, env = NULL, CA_env, model = "full", alpha = 0.05, scale = TRUE)
```

### GDM with model selection

Now, let's run a GDM using only significant variables (i.e., a "best" model). In this case, a variable selection process is done

TODO: describe var sel
TODO: mention that default of this is to NOT include geo dists? It is important to be aware that GDM always considers geographic distance as a variable XXX HOW IS IT CODED

```{r best vars, fig.height=5, fig.width=20, cache = TRUE}
mod_best <- gdm_do_everything(gendist, liz_coords, env = NULL, CA_env, model = "best", nPerm = 10, scale = TRUE)
```

### Results of GDM

There are three elements within the results of `gdm_do_everything()`: "model" (final model), "predictors" (coefficients for predictor variables), and "rast" (PCA RGB map). First, we'll want to see the coefficients from the models obtained.

Let's assume we want to move forward with examining the results from the model selection output. There should be coefficients for each of the significant predictor variables:

```{r preds}
mod_full$predictors
```

### Interpreting the results of GDM

Now, let's create a map:

```{r map}
gdm_map(mod, envlayers, coords)
```

Finally, let's plot the isplines for our best model:

```{r isplines, fig.width = 9, fig.height = 3}
gdm_plot_isplines(mod)
```

We will sum isplines for each significant predictor variable to determine the relative contributions of each.

We can also sum all environmental variables together and compare to geographic distance to get the relative variation explained by IBE (enviro vars) and IBD (geo dist).
