---
title: "LFMM_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LFMM_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Latent factor mixed models (LFMM)

```{r setup}
library(algatr)
library("here")
#to install LFMM:
#devtools::install_github("bcm-uga/lfmm")
library("lfmm") 
library("tidyverse")

#libraries for k selection
library("AssocTests")
library("tess3r")

devtools::load_all()
```

Latent factor mixed models (LFMM; [TODO](TODO REFER)) describe


Similar to RDA, LFMM is a genotype-environment association (GEA) method in which mixed models are used to determine loci that are significantly associated with environmental variables. Importantly, LFMM takes into account -- and corrects for -- unobserved variables that may confound results (known as latent factors). A good example of a latent factor that we may want to correct for prior to examining environment-associated loci is population structure.

As with TESS3 (TODO link vignette?), there are a variety of ways to determine the best number of latent factors (also called K values); algatr provides the option to test three of these methods: (1) Tracy Widom test; (2) quick elbow test; and (3) using the clustering algorithm from TESS (i.e., determining latent factors corresponding to some measure of population structure). We will explore all of these, and compare results, in this vignette.

The main function to perform an LFMM analysis is `lfmm_do_everything()`, which primarily uses the `lfmm2()` function [Caye et al. 2019](https://academic.oup.com/mbe/article/36/4/852/5290100) within the LEA package [TODO](TODO). lfmm2 builds off the lfmm package (see [here](https://bcm-uga.github.io/lfmm/index.html) for details), but is computationally faster.

## Read in and process data files

Running `lfmm_do_everything()` requires two input files: a genotype dosage matrix (the `gen` argument) and the environmental values extracted at sampling coordinates (the `env` argument). If you do not have extracted environmental values (as with the test dataset), `rda_do_everything()` will extract values for us. For clarity's sake here, we'll do the conversion from a vcf to dosage matrix using the `vcf_to_dosage()` function. 

```{r, warning = FALSE}
# Load vcf, genetic dist matrix, and coordinates for 53 inds, and three environmental layers for test dataset
load_example()
# Convert vcf to dosage matrix
gen <- vcf_to_dosage(liz_vcf)
# Also, our code assumes that sample IDs from gendist and coords are the same order; be sure to check this before moving forward!
```

# Run LFMM

## Comparing different LFMM methods

```{r, warning = FALSE,  results = FALSE, message = FALSE, warning = FALSE, fig.height=4, fig.width = 6}
results <- lfmm_do_everything(gen, CA_env, lfmm_method = "ridge", K_selection = "tracy.widom")
```


## Comparing different K selection methods

```{r, warning = FALSE,  results=FALSE, warning = FALSE}
results_tw <- lfmm_do_everything(gen, env, K_selection = "tracy.widom")
```

```{r, results = FALSE, warning = FALSE}
results_qe <- lfmm_do_everything(gen, env = env, K_selection = "quick.elbow")
```

```{r, results = FALSE, cache = TRUE, warning = FALSE}
results_tess <- lfmm_do_everything(gen, env, coords, K_selection = "tess", Kvals = 1:20)
```



