---
title: "TESS_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TESS_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# TESS population structure analysis

```{r setup}
library(algatr)
library(here)
library(tidyverse)
library(raster)
library(rgdal)
library(viridis)
library(terra)
# devtools::install_github("bcm-uga/TESS3_encho_sen")
library(tess3r)

# Need LEA library for cross entropy
#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")
#BiocManager::install("LEA")
library(LEA)

# for plotting (from TESS)
source("http://membres-timc.imag.fr/Olivier.Francois/Conversion.R")
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")

# devtools::install_github("cran/automap")
library(automap)
library(spatialEco)

devtools::load_all()
```

TESS is a method to estimate population structure using ancestry coefficient estimates. As with STRUCTURE, one first describes genetic variation by assigning individuals according to numbers of clusters (K values). Given the "best" number of K values, individuals are then given proportions to which they belong to each K. These proportions correspond to ancestry coefficients, and can be interpreted as the proportion of an individual's ancestry belonging to different ancestral groups.

Unlike STRUCTURE, TESS is spatially explicit, taking into account coordinate data; thus, ancestry coefficient estimates incorporate knowledge of the sampling space. TESS3 [Caye et al. 2016](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.12471) was later developed as an extension of the TESS algorithm, with modifications to the underlying statistical clustering algorithm to increase computational speed.

For additional information on the original development and implementation of the algorithm used by TESS, see [François et al. 2006](https://academic.oup.com/genetics/article/174/2/805/6061653?login=false) and [Chen et al. 2007](https://onlinelibrary.wiley.com/doi/10.1111/j.1471-8286.2007.01769.x). Finally, our code primarily uses the tess3r package (see [here](https://bcm-uga.github.io/TESS3_encho_sen/index.html) for documentation).

### Read in and process input data

Running TESS3 requires three data files for input: a genotype dosage matrix (the `gen` argument), coordinates for samples (the `coords` argument), and environmental layers (the `envlayers` argument). We can use a vcf and the `vcf_to_dosage()` function to convert a vcf to a dosage matrix.

```{r data}
# Load vcf, genetic dist matrix, and coordinates for 53 inds, and three environmental layers for test dataset
load_example()
# Our code assumes that the first column is longitude and second is latitude; check this:
head(liz_coords)
# Also, our code assumes that sample IDs from gendist and coords are the same order; be sure to check this before moving forward!
```

```{r dosage}
# Convert vcf to genotype matrix
liz_dosage <- vcf_to_dosage(liz_vcf)
```

### Process environmental data

TESS provides the ability to map ancestry coefficient estimates, and smooths the resulting surface using a method of spatial interpolation called kriging. We have extended the existing kriging of Q values TODO XXX

```{r}
# First, create a RasterLayer for kriging
# We can use one environmental layer (PC1), aggregated (i.e., increased cell size) to increase computational speed
krig_raster <- raster::aggregate(CA_env[[1]], fact = 6)

# If you want to see the difference between the non-aggregated (original) and aggregated rasters:
terra::plot(CA_env[[1]], col = mako(100), axes = FALSE)
terra::plot(krig_raster, col = mako(100), axes = FALSE)
```

## K selection

------------------------------------------------------------------------

algatr allows users with the option to test a range of K values using the `tess_ktest()` function. K selection is often accomplished by minimizing the cross entropy, but often this over-simplifies the data. The `tess_ktest()` function allows users to specify how K selection is done using the `K_selection` argument; users can set the K values manually (`"manual"`) or using an automatic approach described below (`"auto"`). `tess_ktest()` runs TESS for each value of K within the user-specified range, and outputs cross-validation scores for different K values which allows a user to select the "best" K for their dataset.

### Automatic K selection

The default K selection (`"auto"`) within the `tess_ktest()` function is based on not only minimizing cross-entropy scores (which is typically done), but minimizes the slope of the line that connects cross-entropy scores between K values. In this way, the best K is selected once an "elbow" is visible, or when cross-entropy scores begin to plateau and the slope approaches 0. This function makes use of the approach described [here](https://chazhyseni.github.io/NALgen/post/determining_bestk/).

The resulting object contains `tess3_obj`, which provides the cross-validation scores for the different K values that were tested. Let's start with an automatic K-value testing procedure for K values 1-3.

```{r auto test Ks}
tess3_result_auto <- tess_ktest(liz_dosage, liz_coords, Kvals = 1:10, ploidy = 2, K_selection = "auto")
# Get tess object
tess3_obj <- tess3_result$tess3_obj

# get best K and get Qmatrix
qmat <- qmatrix(tess3_obj, K = 4)
```

### Manual K selection

Now, let's run the same range of K values (1-10) but this time specify the manual K selection (`"manual"` argument).

```{r manual test Ks}
tess3_result_man <- tess_ktest(liz_dosage, liz_coords, Kvals = 1:10, ploidy = 2, K_selection = "manual")
```

### Krig Q values

```{r, cache = TRUE, results = FALSE, warning = FALSE}
# krige Qmatrix
# TODO [EAC]: coords needs to have x and y (if long and lat returns error?)
krig_admix <- tess_krig(qmat, liz_coords, krig_raster)
```

## Interpreting results

------------------------------------------------------------------------

The `results` object contains results for the best-supported K value, including:

1.  The value of the best-supported K (2, in this case): `K`

2.  The matrix of individual ancestry coefficients, or Q values: `Qmatrix`

3.  A RasterBrick containing the kriged ancestry coefficient values for mapping: `krig_admix`

4.  Results from the cross-validation analysis of all values of K: `tess_results`

5.  Sampling coordinates: `coords`

6.  The range of K values that were tested: `Kvals`

7.  The RasterLayer upon which ancestry coefficients are mapped: `grid`

```{r, cache = TRUE}
summary(results)
```

## Plotting

------------------------------------------------------------------------

### Building maps in TESS3

Now let's exploring how Q values (ancestry coefficients) are mapped. We'll move forward with the automatic K selection results (so, a K value of TODO XXX).

### Basic ggplot

```{r, fig.width = 5, fig.height = 5}
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ")
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ_poly")
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ", minQ = 0.20)
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ_poly", minQ = 0.20)
```

```{r}
tess_barplot(qmat, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```

## Extended base R plotting

To plot individual layers and have more control over color breaks, alpha levels, legend/color scale settings, etc. use base R plotting functions

```{r, fig.width = 10, fig.height = 4}
par(mfrow = c(1, nlayers(krig_admix)), mar = rep(2,4), oma = rep(1,4))
tess_plot_allK(krig_admix, col_breaks = 20, legend.width = 2)
```

```{r, fig.width = 5, fig.height = 5, warning = FALSE}
par(mfrow = c(2,2), pty = "s", mar = rep(0,4))

tess_plot(krig_admix, plot_method = "maxQ", col_breaks = 20)
tess_plot(krig_admix, plot_method = "allQ", col_breaks = 20, minQ = 0.20, col_alpha = 0.90)
tess_plot(krig_admix, plot_method = "maxQ_poly")
tess_plot(krig_admix, plot_method = "allQ_poly", minQ = 0.20, col_alpha = 0.50)
```

## Plotting with default tess3r package functions

```{r}
barplot(qmat, sort.by.Q = TRUE, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```

NOTE: the original tess3r package uses a different type of kriging than autoKrige, it also kriges and plots in the same function (so there isn't an easy way to get the raster output).

```{r, fig.width = 5, fig.height = 5}
plot(qmat, 
     liz_coords, 
     method = "map.max", 
     interpol = FieldsKrigModel(10),  
     main = "Ancestry coefficients",
     xlab = "Longitude", ylab = "Latitude", 
     col.palette = CreatePalette(),
     resolution = c(300,300), cex = .4)

# comparison to algatr
par(pty = "s", mar = rep(0,4))
tess_plot(krig_admix, liz_coords, plot_method = "maxQ", col_pal =  tess_col_default)
```

## Running TESS with `tess_do_everything()`

------------------------------------------------------------------------

The algatr package also has an option to run all of the above functionality in a single function, `tess_do_everything()`. This function will TODO XXXXX. The resulting object looks identical to the output object from TODO XXXX.  *(N.B.: the `tess_do_everything()` function does not require a dosage matrix; it will do the conversion automatically if a vcf is provided)*.

The `tess_do_everything()` function will output (1) a plot with cross-entropy scores for specified K values with a vertical dashed line indicating the best-supported K value; (2) a map with ancestry coefficients plotted and colorized based on corresponding Q values for the best-supported K; and (3) a standard STRUCTURE-style bar plot in which vertical bars represent individuals and the proportion of color of the stacked bars corresponds to the ancestry coefficients of each individual (i.e,. their assignment into each K) for the best-supported K.

```{r do everything, cache = TRUE, results = FALSE, warning = FALSE, fig.width = 5, fig.height = 5}
# TODO [EAC]: autoKrige duplicate observations in input_data?
# TODO [EAC]: printed order?
par(pty = "s", mar = rep(2,4))
# results <- tess_do_everything(gen, liz_coords, CA_env[[1]], Kvals = 1:3, K_selection = "auto", minQ = 0.10)
results_krige <- tess_do_everything(gen, liz_coords, krig_raster, Kvals = 1:3, K_selection = "auto", minQ = 0.10)
```
