---
title: "TESS_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TESS_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# TESS population structure analysis

```{r setup}
library(algatr)
library(here)
library(tidyverse)
library(raster)
library(rgdal)
# Figure out libraries later
# to install TESS:
# install.packages("devtools")
# devtools::install_github("bcm-uga/TESS3_encho_sen")
library(tess3r)
```

# for plotting (from TESS)
source("http://membres-timc.imag.fr/Olivier.Francois/Conversion.R")
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")

# need LEA library for cross entropy
#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")
#BiocManager::install("LEA")
#library("LEA")

# devtools::install_github("cran/automap")
library(automap)
library(spatialEco)

devtools::load_all()
```
## Load Test Data

```{r, warning = FALSE}
load_example()
```

```{r}
# convert vcf to genotype matrix
gen <- vcf_to_dosage(liz_vcf)
```

```{r}
# create a RasterLayer for kriging
# here we just use one environmental layer, aggregated to increase computational speed
krig_raster <- raster::aggregate(CA_env[[1]], 4)

# alternatively a raster can be created from coordinates
krig_raster2 <- coords_to_raster(liz_coords, res = 0.01, buffer = 1, plot = TRUE)
```

## Run everything


```{r, cache = TRUE, results = FALSE, warning = FALSE, fig.width = 5, fig.height = 5}
par(pty = "s", mar = rep(2,4))
results <- tess_do_everything(gen, liz_coords, CA_env[[1]], Kvals = 1:10, K_selection = "auto", minQ = 0.10)
```


## Line by line TESS

### Run TESS

```{r, cache = TRUE}
# run tess with multiple K values
tess3_result <- tess_ktest(gen, liz_coords, Kvals = 1:10, ploidy = 2) 

# look at resulting object (ANNE: provide a description of what is in this object)
summary(tess3_result)

# get tess object
tess3_obj <- tess3_result$tess3_obj

# or run with one K value
tess3_obj <- tess3(gen, coord = as.matrix(liz_coords), K = 3, ploidy = 2) 

# get best K and get Qmatrix
qmat <- qmatrix(tess3_obj, K = 4)

```

### Krig Q values

```{r, cache = TRUE, results = FALSE, warning = FALSE}
# krige Qmatrix
# TODO [EAC]: coords needs to have x and y (if long and lat returns error?)
krig_admix <- tess_krig(qmat, liz_coords, krig_raster)
```

## Plotting 

### Basic ggplot

```{r, fig.width = 5, fig.height = 5}
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ")
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ_poly")
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ", minQ = 0.20)
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ_poly", minQ = 0.20)
```

```{r}
tess_barplot(qmat, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```

## Extended base R plotting

To plot individual layers and have more control over color breaks, alpha levels, legend/color scale settings, etc. use base R plotting functions

```{r, fig.width = 10, fig.height = 4}
par(mfrow = c(1, nlayers(krig_admix)), mar = rep(2,4), oma = rep(1,4))
tess_plot_allK(krig_admix, col_breaks = 20, legend.width = 2)
```

```{r, fig.width = 5, fig.height = 5, warning = FALSE}
par(mfrow = c(2,2), pty = "s", mar = rep(0,4))

tess_plot(krig_admix, plot_method = "maxQ", col_breaks = 20)
tess_plot(krig_admix, plot_method = "allQ", col_breaks = 20, minQ = 0.20, col_alpha = 0.90)
tess_plot(krig_admix, plot_method = "maxQ_poly")
tess_plot(krig_admix, plot_method = "allQ_poly", minQ = 0.20, col_alpha = 0.50)
```

## Plotting with default tess3r package functions

```{r}
barplot(qmat, sort.by.Q = TRUE, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```


NOTE: the original tess3r package uses a different type of kriging than autoKrige, it also kriges and plots in the same function (so there isn't an easy way to get the raster output). 
```{r, fig.width = 5, fig.height = 5}
plot(qmat, 
     liz_coords, 
     method = "map.max", 
     interpol = FieldsKrigModel(10),  
     main = "Ancestry coefficients",
     xlab = "Longitude", ylab = "Latitude", 
     col.palette = CreatePalette(),
     resolution = c(300,300), cex = .4)

# comparison to algatr
par(pty = "s", mar = rep(0,4))
tess_plot(krig_admix, liz_coords, plot_method = "maxQ", col_pal =  tess_col_default)
```
