---
title: "TESS_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TESS_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# TESS population structure analysis

```{r setup}
library(algatr)
library(here)
library(tidyverse)
library(raster)
library(rgdal)
library(viridis)
library(terra)
# devtools::install_github("bcm-uga/TESS3_encho_sen")
library(tess3r)

# Need LEA library for cross entropy
#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")
#BiocManager::install("LEA")
library(LEA)

# for plotting (from TESS)
source("http://membres-timc.imag.fr/Olivier.Francois/Conversion.R")
source("http://membres-timc.imag.fr/Olivier.Francois/POPSutilities.R")

# devtools::install_github("cran/automap")
library(automap)
library(spatialEco)

devtools::load_all()
```

TESS is a method to estimate population structure using ancestry coefficient estimates. As with STRUCTURE, one first describes genetic variation by assigning individuals according to numbers of clusters (K values). Given the "best" number of K values, individuals are then given proportions to which they belong to each K. These proportions correspond to ancestry coefficients, and can be interpreted as the proportion of an individual's ancestry belonging to different ancestral groups.

Unlike STRUCTURE, TESS is spatially explicit, taking into account coordinate data; thus, ancestry coefficient estimates incorporate knowledge of the sampling space. TESS3 [Caye et al. 2016](https://onlinelibrary.wiley.com/doi/10.1111/1755-0998.12471) was later developed as an extension of the TESS algorithm, with modifications to the underlying statistical clustering algorithm to increase computational speed.

Researchers can determine K values in a variety of ways, and here we provide TODO [EAC] XXX ways to do so: XXX.

For additional information on the original development and implementation of the algorithm used by TESS, see [François et al. 2006](https://academic.oup.com/genetics/article/174/2/805/6061653?login=false) and [Chen et al. 2007](https://onlinelibrary.wiley.com/doi/10.1111/j.1471-8286.2007.01769.x). Finally, our code primarily uses the tess3r package (see [here](https://bcm-uga.github.io/TESS3_encho_sen/index.html) for documentation).

### Read in and process input data

Running TESS3 requires three data files for input: a genotype dosage matrix (the `gen` argument), coordinates for samples (the `coords` argument), and environmental layers (the `envlayers` argument). We can use a vcf and the `vcf_to_dosage()` function to convert a vcf to a dosage matrix. *(N.B.: the `tess_do_everything()` function does not require a dosage matrix; it will do the conversion automatically if a vcf is provided)*.

```{r data}
# Load vcf, genetic dist matrix, and coordinates for 53 inds, and three environmental layers for test dataset
load_example()
# Our code assumes that the first column is longitude and second is latitude; check this:
head(liz_coords)
# Also, our code assumes that sample IDs from gendist and coords are the same order; be sure to check this before moving forward!
```

```{r}
# Convert vcf to genotype matrix
gen <- vcf_to_dosage(liz_vcf)
```

## Process environmental data

TESS3 provides the ability to map ancestry coefficient estimates, and smooths the resulting surface using a method of spatial interpolation called kriging. We have extended the existing kriging of Q values TODO XXX

```{r}
# First, create a RasterLayer for kriging
# We can use one environmental layer (PC1), aggregated (i.e., increased cell size) to increase computational speed
krig_raster <- raster::aggregate(CA_env[[1]], fact = 4)

# If you want to see the difference between the non-aggregated (original) and aggregated rasters:
terra::plot(CA_env[[1]], col = mako(100), axes = FALSE)
terra::plot(krig_raster, col = mako(100), axes = FALSE)

# Alternatively, a raster can be created from coordinates using the `coords_to_raster()` function
krig_raster_coords <- coords_to_raster(liz_coords, res = 0.01, buffer = 1, plot = TRUE)
```

### Run TESS3

```{r, cache = TRUE, results = FALSE, warning = FALSE, fig.width = 5, fig.height = 5}
par(pty = "s", mar = rep(2,4))
results <- tess_do_everything(gen, liz_coords, CA_env[[1]], Kvals = 1:3, K_selection = "auto", minQ = 0.10)
```

## Line by line TESS

### Run TESS

```{r, cache = TRUE}
# run tess with multiple K values
tess3_result <- tess_ktest(gen, liz_coords, Kvals = 1:10, ploidy = 2) 

# look at resulting object (ANNE: provide a description of what is in this object)
summary(tess3_result)

# get tess object
tess3_obj <- tess3_result$tess3_obj

# or run with one K value
tess3_obj <- tess3(gen, coord = as.matrix(liz_coords), K = 3, ploidy = 2) 

# get best K and get Qmatrix
qmat <- qmatrix(tess3_obj, K = 4)

```

### Krig Q values

```{r, cache = TRUE, results = FALSE, warning = FALSE}
# krige Qmatrix
# TODO [EAC]: coords needs to have x and y (if long and lat returns error?)
krig_admix <- tess_krig(qmat, liz_coords, krig_raster)
```

## Plotting

### Basic ggplot

```{r, fig.width = 5, fig.height = 5}
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ")
tess_ggplot(krig_admix, liz_coords, plot_method = "maxQ_poly")
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ", minQ = 0.20)
tess_ggplot(krig_admix, liz_coords, plot_method = "allQ_poly", minQ = 0.20)
```

```{r}
tess_barplot(qmat, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```

## Extended base R plotting

To plot individual layers and have more control over color breaks, alpha levels, legend/color scale settings, etc. use base R plotting functions

```{r, fig.width = 10, fig.height = 4}
par(mfrow = c(1, nlayers(krig_admix)), mar = rep(2,4), oma = rep(1,4))
tess_plot_allK(krig_admix, col_breaks = 20, legend.width = 2)
```

```{r, fig.width = 5, fig.height = 5, warning = FALSE}
par(mfrow = c(2,2), pty = "s", mar = rep(0,4))

tess_plot(krig_admix, plot_method = "maxQ", col_breaks = 20)
tess_plot(krig_admix, plot_method = "allQ", col_breaks = 20, minQ = 0.20, col_alpha = 0.90)
tess_plot(krig_admix, plot_method = "maxQ_poly")
tess_plot(krig_admix, plot_method = "allQ_poly", minQ = 0.20, col_alpha = 0.50)
```

## Plotting with default tess3r package functions

```{r}
barplot(qmat, sort.by.Q = TRUE, border = NA, space = 0, xlab = "Individuals", ylab = "Ancestry coefficients")
```

NOTE: the original tess3r package uses a different type of kriging than autoKrige, it also kriges and plots in the same function (so there isn't an easy way to get the raster output).

```{r, fig.width = 5, fig.height = 5}
plot(qmat, 
     liz_coords, 
     method = "map.max", 
     interpol = FieldsKrigModel(10),  
     main = "Ancestry coefficients",
     xlab = "Longitude", ylab = "Latitude", 
     col.palette = CreatePalette(),
     resolution = c(300,300), cex = .4)

# comparison to algatr
par(pty = "s", mar = rep(0,4))
tess_plot(krig_admix, liz_coords, plot_method = "maxQ", col_pal =  tess_col_default)
```
